Limits to Ecological Forecasting: Estimating Uncertainty for Critical Transitions with Deep Learning

### 2022-09-20 manuscript/manuscript.Rmd
Read through finished

We provide fully reproducible coded examples in R and Python for performing, scoring, and visualizing each of the forecasts considered here.
After significant time spent considering alternative frameworks, we have emphasized those which best met our requirements for performance, ease-of-use, flexibility, and support for the latest probabilistic machine learning models for forecasting. 
Most of our forecasts use the `darts` framework, a sophisticated and well documented Python library with support for a wide range of methods.
Our model-based MCMC forecasts use the `greta` framework, a R library that uses Python-based `tensorflow` probability to achieve better performance.
While Python-based frameworks currently have the edge in performance and access modern ML algorithms, they lag behind in attention to statistical issues such as the computation of strictly proper skill scores.

@@

The Nicholson-Bailey model describes a predator-prey dynamic for the relationship of a host species and an obligate parasitoid, originally used to model the population dynamics of blowflies (_Lucilia cuprina_) [@nicholson_balance_1935; @nicholson_outline_1954; @nicholson_compensatory_1954].
We consider the form which includes density dependence in the host species, and allow for environmental stochasticity,

@@

Where $H_t$ is the population density of the host species at time $t$, (in arbitrary units) and $P_t$ the population density of the parasitoid.
The time step is defined by the generation time of the parasitoid, which is about two weeks in the case of Nicholson's blowflies [@nicholson_outline_1954].
Following @dakos_methods_2012, we further allow the carrying capacity of the host, $K$ to slowly increase at a linear rate, which drives a supercritical Hopf bifurcation as $K$ becomes sufficiently large. 
In a Hopf bifurcation, a stable node starts an oscillatory pattern which grows in amplitude as the bifurcation parameter continues to increase.
In this model, the Hopf bifurcation is dubbed 'supercritical' as it creates a stable limit cycle instead of an unstable one.
This example illustrates one of the many kinds of challenges which nonlinear phenomena pose to forecasting: the "historical" data prior to the bifurcation never exhibit the cyclical dynamics of growing amplitude that will emerge after the bifurcation occurs.
If we had used a purely deterministic model, the dynamics would be constrained to a single stable point, corresponding to a slowly changing steady-state population size of host and parasitoid populations.
However, stochasticity in this case acts as a source of some additional information about the dynamics, as the noise excites quasi-cycles which are visible in the irregular oscillations that appear significantly prior to the emergence of true limit cycles which follow the bifurcation [@boettiger_noise_2018].
Examples use the following parameters: $H_0 = 9$, $P_0 = 1$, $r = 0.75$, $c = 0.1$, $K_0 = 14$, $\delta = 0.08$, $\sigma_H = 0.02$, $\sigma_P = 0.02$.

@@

## Selecting timescales
In each scenario,  t = 0 is the start time of the training data, while the length of training data and forecast horizon (with ensembles sampled from the true distribution) is illustrated in Fig 1. 
For the Hopf bifurcation, forecast begins at t=100 and extends to t =200, for the saddle node, forecast begins at t=250 and extends to t = 500, for the stochastic transition, both training data and forecasting tasks begin at 0 and extend to t = 250. 
While much attention is often paid to the number of data points in training or testing data, it is essential to realize that these are only meaningful relative to the specific process in question. 
Thus, in each case we have selected these time intervals to focus on the dynamical process in question, which unfolds at a different rate and tempo in each scenario. 
For instance, if the stochastic scenario was restricted to a much shorter timescale used in the Hopf case, few replicate simulations would experience a transition at all. 
If it were made much longer, most of the timeseries would be spent post-transition. 
Likewise, if the forecast for the Hopf scenario was extended much further into the future under the current parameterization, the system experiences a homoclinic bifurcation at which the population collapses to 0. 
Using different length timescales allows us to consider the three different forecasting tasks illustrated in Fig 1 that focus around predicting the critical behavior, rather than predicting long periods of relative stasis.
These three critical transitions are fundamentally different processes, there is no perfect apples-to-apples parameterization for each that allows the transition to unfold in a way that gives precisely the same time windows.

@@

Recurrent neural networks are the predominant memory-based deep learning method.
Recurrent neural networks differ from feed-forward neural networks in that a recurrent neural network provides feedback to itself between time steps [@sherstinsky_fundamentals_2020].
By providing self-feedback, recurrent neural networks are able to retain information from previous time steps and thus learn temporal dependencies.
However, a standard recurrent neural network is unwieldy to train because of the vanishing and exploding gradient problem [@pascanu_difficulty_2013], so there have been specialized neural network architectures designed to avoid these gradient problems.
Long Short-term Memory and Gated Recurrent Units Networks are considered to be the state of the art recurrent neural networks that address exploding and vanishing gradients [@chung_empirical_2014].
These methods avoid gradient problems by regulating the self-feedback via gates which perform operations on the feedback signal -- see @chung_empirical_2014 for more details.
While GRU's and LSTM's commonly outperform standard RNN's, it is difficult to anticipate whether GRU's or LSTM's will be best suited for any time series problem [@chung_empirical_2014], so we investigate both methods.

@@

### Encoder-Decoder

Encoder-decoder architectures have been shown empirically to excel on sequence to sequence tasks [@aitken_understanding_2021].
Encoder-decoders work by processing the input sequence into a fixed-length vector then decoding the fixed-length vector to the predicted output sequence.
It is thought that by encoding the input sequence to a vector, encoder-decoders find informative representations of the input sequence that make the prediction task much easier [@sutskever_sequence_2014].
Note that it is possible to use any type of neural network as the encoder and the decoder, but it is most common to use recurrent neural networks or networks with attention mechanisms [@aitken_understanding_2021].
Of the models that we present, Block RNNs and Transformers have encoder-decoder-based architectures.

@@

To compare forecasts, we focus exclusively on metrics of forecast skill which satisfy the property from @gneiting_strictly_2007 of a strictly proper score.
This ensures the very desirable behavior that no probabilistic forecast $Q(x,t)$ can have a score as high as the score of the true process $P(x,t)$ on average. 
In other words, while it is possible for any of the models considered to _overfit_ the data against which they are trained, i.e. have a higher likelihood than the true process, it is not possible for these models to overfit the data against which they are scored.
It is worth noting that this property applies specifically to probabilistic forecasts and not point forecasts. Not all common metrics often used to compare forecasts are strictly proper -- such as the average root-mean-square error or the average absolute error. <!-- MFL: hmm I am not completely following this point here. CB: added some examples of non-strictly-proper scoring metrics. For instance, {darts} has a whole 'metrics' class, which does not include any strictly proper metrics (because ML researchers don't care about statistical theorems?) -->
Concerns about over-fitting arise in most types of model estimation and are a particularly acute concern to machine learning methods due to the bias-variance trade-off [@mehta_high-bias_2019]. <!-- MFL: I don't think I nailed the overfitting part in ML section but we probably want to reference this here-->
This makes the use of strictly proper scoring especially relevant in assessing machine learning predictions.

@@

Forecasts of the Hopf bifurcation (Fig 3) are roughly comparable across the phenomenological models (ARIMA and machine learning models).
All models are trained using 100 time points drawn from the period of time prior to the onset of the Hopf bifurcation, which leads to a stable limit cycle that gradually grows in magnitude.
Most models predict a roughly constant mean with a spread roughly equal to that created by the stochastic oscillations around the stable node as seen in the training data prior to the bifurcation.
Notably, the GRU method picks up the oscillatory nature of the dynamics, despite the fact that no true oscillations were yet present in the training data.
However, like the other ML models, it fails to predict the growing amplitude of those oscillations.
Having access to the true model structure, the MCMC model alone predicts the transition into a pattern of oscillations which grows over time, though it tends to overestimate the amplitude of those oscillations initially.
Despite this, overall all methods score comparably in CRPS score (Fig 2) with most ML methods actually out-performing the MCMC score on average (Fig 5), albeit with much greater variation in individual scores.
A clearer picture can be seen by looking at these skill scores over time (Fig 6-7), which show that MCMC is initially performing worse (over-predicting variance) but as oscillations grow further, it starts outperforming the more stationary forecasts of the ML models.

@@

The saddle node bifurcation proves even more difficult for most methods (Fig 4).
Only the MCMC model anticipates the sharp transition to an alternative state, though this behavior is more baked into the method from the start which takes the true model structure as given. 
Even accurate estimation of the MCMC requires slightly informative priors, though still broad enough to reflect a wide range of possible outcomes. 
Two ML methods -- Block RNNs and Transformers -- resemble a naive prediction extrapolating the last observed state, failing even to reflect the slow downward trend of the training data. 
LSTM indicates greater uncertainty, while GRU shows very large variability which spans the alternative stable state range. 
With additional tuning, better performance may be possible for these ML models. 
The selected ARIMA model reflects wide uncertainty that is nevertheless not broad enough to span the alternative stable state.
Consequentially, the MCMC estimate easily outperforms the ML models (Fig 2).

@@

Examining patterns in the scores over time (Fig 6-7) provides a more nuanced understanding of the forecast dynamics than aggregate scores alone (Fig 2).
In the Hopf bifurcation, CRPS scores get worse over time across all methods, including the MCMC forecasts.
In the saddle node bifurcation and stochastic transition, the same pattern holds somewhat more dramatically for non-MCMC forecast, while MCMC scores are worst in mid-range.
Comparing CRPS scores to logs score also emphasizes the relative role of uncertainty: for instance, the MCMC scores for the Hopf bifurcation get steadily worse under CRPS but not under logs score. 
A relatively sharp transition can be seen under both MCMC scores on the Hopf bifurcation once the magnitude of the oscillations exceeds the variance created by mere stochasticity: the MCMC model no longer over-estimates the spread of the data, while the ML models now underestimate that variation. 
CRPS scores for stochastic transitions exhibit a distinct two-branch pattern, with scores for a given replicate being either very high (poor skill) or very low, reflecting whether the individual 'true' replicate matches the mean state predicted by the forecast or the other state. 
Logs skill score may be a better measure in this context, where correctly capturing the uncertainty in the forecast means that this bi-modal structure in scores can be avoided entirely, e.g. by the MCMC predictions. 
The forecast-skill-over time plots illustrate different reasons for the bi-modal distribution in skill seen for the saddle-node and stochastic transition scenarios in Fig 2 respectively: in the case of the saddle node, the two modes are distinguished by time-horizon: short term forecasts are relatively accurate, and longer term forecasts (i.e. after the catastrophic transition) are poor.
In the case of the stochastic transition model, the two modes are not structured by horizon but by replicate, with some replicates having transitioned and others still in the original state.

@@

The MCMC case illustrates some of the hard limits to ecological forecasting of critical transitions.
Our MCMC forecast assumes the _true_ model structure, the data-generating process, is known, and the forecaster need only infer the posterior distribution of model parameters.
This is a much stronger assumption than that made by the ML models, though this assumption can potentially be justified on the basis of a mechanistic understanding of the processes involved.
It is important not to confound the MCMC example here with the use of MCMC in process based models of real systems.
In the real world, this is never the case: all models are at best approximations of the underlying processes [@Oreskes1994]. <!-- ML: maybe "In the real world, having access to the _true_ model is never the case:"-->
Despite this advantage, even the MCMC forecasts differ from the distribution of the true process.
Because the available data come from only a small region of the dynamical state space, they are consistent with many possible parameterizations of the same model structure -- which creates likelihood ridges and non-identifiability of specific parameter values.
Using more simplified versions of the dynamical processes in question, such as the canonical form of a bifurcation, can mitigate this issue in some cases.
Even when such non-identifiability issues cannot be avoided entirely, they can usually be diagnosed by examining the degree of mixing in MCMC sampling and comparing posterior to prior distributions.
<!-- While similar convergence issues may arise in the estimation of ML models, they may be harder to diagnose. -->

@@

There are a number of questions that we have left unanswered that we hope will be addressed in future work.
In this paper, we have explored a small number of machine learning and statistical models that can be used for forecasting, so comprehensive conclusions can not be drawn on whether statistical or machine learning-based approaches are better suited for critical transition forecasting problems.
Neither can we claim that the ML methods employed will translate well to all sudden transition event forecasting problems in reality, since working with real data will introduce additional difficulties like how to deal with missing data, sparse data and observation errors.

Furthermore, our analysis has focused on the task of making a single forecast prior to the occurrence of a critical transition.
Forecasting is ideally a more iterative process of data assimilation, where forecasts are updated with respect to additional observations, rather than projecting 100s of time steps into the future [@dietze_iterative_2018]. 
Updating a forecast after a critical transition has already occurred may be of little use in the context of hysteresis, such as under the saddle node or stochastic transition -- recognizing the alternative stable state only after the system is stuck in that basin will often be considered 'too late'.
Assimilation may be more applicable to the Hopf bifurcation, where additional observations of slowly growing oscillations may lead to more accurate forecasts. 
Such models may even accurately predict the homoclinic bifurcation that occurs when the limit cycle grows too large, eventually hitting a saddle point of zero population size for the host species.
We leave these cases to future exploration rather than attempting to explore all such variations in a single narrative. 
We have focused on these illustrative examples accompanied by an extensive appendix of reproducible code implemented using well documented and intuitive frameworks.
We encourage the reader to use these appendices as an entry point into further exploration.

>>>>

We provide fully reproducible coded examples in R and Python for performing, scoring, and visualizing each of the forecasts considered here.
After significant time spent considering alternative frameworks, we have emphasized those which best met our requirements for performance, ease-of-use, flexibility, and support for the latest probabilistic machine learning models for forecasting. 
Most of our forecasts use the `darts` framework, a sophisticated and well documented Python library with support for a wide range of methods.
Our model-based MCMC forecasts use the `greta` framework, a R library that uses Python-based `TensorFlow Probability` to achieve better performance.
While Python-based frameworks currently have the edge in performance and access modern ML algorithms, they lag behind in attention to statistical issues such as the computation of strictly proper skill scores.

@@

The Nicholson-Bailey model describes a predator-prey dynamic for the relationship of a host species and an obligate parasitoid, originally used to model the population dynamics of blowflies (_Lucilia cuprina_) [@nicholson_balance_1935; @nicholson_outline_1954; @nicholson_compensatory_1954].
We consider the form which includes density dependence in the host species, and we allow for environmental stochasticity,

@@

Where $H_t$ is the population density of the host species at time $t$, (in arbitrary units) and $P_t$ is the population density of the parasitoid.
The time step is defined by the generation time of the parasitoid, which is about two weeks in the case of Nicholson's blowflies [@nicholson_outline_1954].
Following @dakos_methods_2012, we further allow the carrying capacity of the host, $K$ to slowly increase at a linear rate, which drives a supercritical Hopf bifurcation as $K$ becomes sufficiently large. 
In a Hopf bifurcation, a stable node starts an oscillatory pattern which grows in amplitude as the bifurcation parameter continues to increase.
In this model, the Hopf bifurcation is dubbed 'supercritical' as it creates a stable limit cycle instead of an unstable one.
This example illustrates one of the many kinds of challenges which nonlinear phenomena pose to forecasting: the "historical" data prior to the bifurcation never exhibit the cyclical dynamics of growing amplitude that will emerge after the bifurcation occurs.
If we had used a purely deterministic model, the dynamics would be constrained to a single stable point, corresponding to a slowly changing steady-state population size of host and parasitoid populations.
However, stochasticity in this case acts as a source of some additional information about the dynamics, as the noise excites quasi-cycles which are visible in the irregular oscillations that appear significantly prior to the emergence of true limit cycles which follow the bifurcation [@boettiger_noise_2018].
Examples use the following parameters: $H_0 = 9$, $P_0 = 1$, $r = 0.75$, $c = 0.1$, $K_0 = 14$, $\delta = 0.08$, $\sigma_H = 0.02$, $\sigma_P = 0.02$.

@@

## Selecting timescales
In each scenario,  t=0 is the start time of the training data, while the length of training data and forecast horizon (with ensembles sampled from the true distribution) are illustrated in Fig 1. 
For the Hopf bifurcation, forecasts begin at t=100 and extend to t=200; for the saddle node, forecasts begin at t=250 and extend to t=500; and, for the stochastic transition, both training data and forecasting tasks begin at 0 and extend to t=250. 
While much attention is often paid to the number of data points in training or testing data, it is essential to realize that these are only meaningful relative to the specific process in question. 
Thus, in each case, we have selected these time intervals to focus on the dynamical process in question, which unfolds at a different rate and tempo in each scenario.
For instance, if the stochastic scenario was restricted to the much shorter timescale used in the Hopf case, few replicate simulations would experience a transition at all. 
If length of the stochastic transition timeseries was made much longer, most of the timeseries would be spent post-transition. 
Likewise, if the forecast horizon for the Hopf scenario was extended much further into the future under the current parameterization, the system would experience a homoclinic bifurcation at which the population collapses to 0. 
Using different length timescales allows us to consider the three different forecasting tasks illustrated in Fig 1 that focus around predicting the critical behavior, rather than predicting long periods of relative stasis.
These three critical transitions are fundamentally different processes, there is no perfect apples-to-apples parameterization for each that allows the transition to unfold in a way that gives precisely the same time windows.

@@

Recurrent neural networks (RNN's) are the predominant memory-based deep learning method.
Recurrent neural networks differ from feed-forward neural networks in that a recurrent neural network provides feedback to itself between time steps [@sherstinsky_fundamentals_2020].
By providing self-feedback, recurrent neural networks are able to retain information from previous time steps and thus learn temporal dependencies.
However, a standard recurrent neural network is unwieldy to train because of the vanishing and exploding gradient problem [@pascanu_difficulty_2013], so there have been specialized neural network architectures designed to avoid these gradient problems.
Long Short-term Memory (LSTM) and Gated Recurrent Units (GRU) Networks are considered to be the state of the art recurrent neural networks that address exploding and vanishing gradients [@chung_empirical_2014].
These methods avoid gradient problems by regulating the self-feedback via gates which perform operations on the feedback signal -- see @chung_empirical_2014 for more details.
While GRU's and LSTM's commonly outperform standard RNN's, it is difficult to anticipate whether GRU's or LSTM's will be best suited for any time series problem [@chung_empirical_2014], so we investigate both methods.

@@

### Encoder-Decoders

Encoder-decoder architectures have been shown empirically to excel on sequence to sequence tasks [@aitken_understanding_2021].
Encoder-decoders work by processing the input sequence into a fixed-length vector then decoding this fixed-length vector to an output sequence.
It is thought that by encoding the input sequence to a vector, encoder-decoders find informative representations of the input sequence that make the prediction task much easier [@sutskever_sequence_2014].
Note that it is possible to use any type of neural network as the encoder and the decoder, but it is most common to use recurrent neural networks or networks with attention mechanisms [@aitken_understanding_2021].
Of the models that we present, Block RNN's are a direct example of an encoder-decoder-based model since a Block RNN employs a RNN as an encoder and a separate RNN as a decoder.
Transformers also have an encoder and decoder component.

@@

To compare forecasts, we focus exclusively on metrics of forecast skill which satisfy the property from @gneiting_strictly_2007 of a strictly proper score.
This ensures the very desirable behavior that no probabilistic forecast $Q(x,t)$ can have a score as high as the score of the true process $P(x,t)$ on average. 
In other words, while it is possible for any of the models considered to _overfit_ the data against which they are trained, i.e. have a higher likelihood than the true process, it is not possible for these models to overfit the data against which they are scored.
It is worth noting that this property applies specifically to probabilistic forecasts and not point forecasts. Not all common metrics often used to compare forecasts are strictly proper -- such as the average root-mean-square error or the average absolute error.
Concerns about over-fitting arise in most types of model estimation and are a particularly acute concern to machine learning methods due to the bias-variance trade-off [@mehta_high-bias_2019].
This makes the use of strictly proper scoring especially relevant in assessing machine learning predictions.

@@

Forecasts of the Hopf bifurcation (Fig 3) are roughly comparable across the phenomenological models (ARIMA and machine learning models).
All models are trained using 100 time points drawn from the period of time prior to the onset of the Hopf bifurcation, which leads to a stable limit cycle that gradually grows in magnitude.
Most models predict a roughly constant mean with a spread roughly equal to that created by the stochastic oscillations around the stable node as seen in the training data prior to the bifurcation.
Notably, the GRU model picks up the oscillatory nature of the dynamics, despite the fact that no true oscillations were yet present in the training data.
However, like the other ML models, it fails to predict the growing amplitude of those oscillations.
Having access to the true model structure, the MCMC model alone predicts the transition into a pattern of oscillations which grows over time, though it tends to overestimate the amplitude of those oscillations initially.
Despite this, all methods score comparably in CRPS score (Fig 2) with most ML methods actually out-performing the MCMC score on average (Fig 5), albeit with much greater variation in individual scores.
A clearer picture can be seen by looking at these skill scores over time (Fig 6-7), which show that MCMC is initially performing worse (over-predicting variance) but as oscillations grow further, it starts outperforming the more stationary forecasts of the ML models.

@@

The saddle node bifurcation proves even more difficult for most methods (Fig 4).
Only the MCMC model anticipates the sharp transition to an alternative state.
Even accurate estimation of the MCMC requires slightly informative priors, though still broad enough to reflect a wide range of possible outcomes. 
Two ML methods -- Block RNN's and Transformers -- resemble a naive prediction extrapolating the last observed state, failing to reflect the slow downward trend of the training data. 
LSTM's indicate greater uncertainty, while GRU's show very large variability which spans the alternative stable state range. 
With additional tuning, better performance may be possible for these ML models. 
The selected ARIMA model reflects wide uncertainty that is nevertheless not broad enough to span the alternative stable state.
Consequentially, the MCMC estimate easily outperforms the ML models (Fig 2).

@@

Examining patterns in the scores over time (Fig 6-7) provides a more nuanced understanding of the forecast dynamics than aggregate scores alone (Fig 2).
In the Hopf bifurcation, CRPS scores get worse over time across all methods, including the MCMC forecasts.
In the saddle node bifurcation and stochastic transition, the same pattern holds somewhat more dramatically for non-MCMC forecasts, while MCMC scores are at their worst around the middle of the forecast horizon.
Comparing CRPS scores to logs score also emphasizes the relative role of uncertainty: for instance, the MCMC scores for the Hopf bifurcation get steadily worse under CRPS but not under logs score. 
A relatively sharp transition can be seen under both MCMC scores on the Hopf bifurcation once the magnitude of the oscillations exceeds the variance created by mere stochasticity: the MCMC model no longer over-estimates the spread of the data, while the ML models now underestimate that variation. 
CRPS scores for stochastic transitions exhibit a distinct two-branch pattern, with scores for a given replicate being either very high (poor skill) or very low, reflecting whether the individual 'true' replicate matches the mean state predicted by the forecast. 
Logs skill score may be a better measure in this context, where correctly capturing the uncertainty in the forecast means that this bi-modal structure in scores can be avoided entirely, e.g. by the MCMC predictions. 
The forecast-skill-over time plots illustrate different reasons for the bi-modal distribution in skill seen for the saddle-node and stochastic transition scenarios in Fig 2 respectively: in the case of the saddle node, the two modes are distinguished by time-horizon; short term forecasts are relatively accurate, and longer term forecasts (i.e. after the catastrophic transition) are poor.
In the case of the stochastic transition model, the two modes are not structured by horizon but by replicate, with some replicates having transitioned and others still in the original state.

@@

The MCMC case illustrates some of the hard limits to ecological forecasting of critical transitions.
Our MCMC forecast assumes that the data-generating process is known, so the forecaster need only infer the posterior distribution of model parameters.
This is a much stronger assumption than that made by the ML models, though this assumption can potentially be justified on the basis of a mechanistic understanding of the processes involved.
It is important not to confound the MCMC example here with the use of MCMC in process based models of real systems.
In the real world, this is never the case: all models are at best approximations of the underlying processes [@Oreskes1994].
Despite this advantage, even the MCMC forecasts differ from the distribution of the true process.
Because the available data come from only a small region of the dynamical state space, they are consistent with many possible parameterizations of the same model structure -- which creates likelihood ridges and non-identifiability of specific parameter values.
Using more simplified versions of the dynamical processes in question, such as the canonical form of a bifurcation, can mitigate this issue in some cases.
Even when such non-identifiability issues cannot be avoided entirely, they can usually be diagnosed by examining the degree of mixing in MCMC sampling and comparing posterior to prior distributions.
<!-- While similar convergence issues may arise in the estimation of ML models, they may be harder to diagnose. -->

@@

There are a number of questions that we have left unanswered that we hope will be addressed in future work.
In this paper, we have explored a small number of machine learning and statistical models that can be used for forecasting, so comprehensive conclusions can not be drawn on whether statistical or machine learning-based approaches are better suited for critical transition forecasting problems.
Neither can we claim that ML methods will translate well to all sudden transition event forecasting problems in reality, since working with real data will introduce additional difficulties like how to deal with missing data, sparse data and observation errors.

Furthermore, our analysis has focused on the task of making a single forecast prior to the occurrence of a critical transition.
Forecasting is ideally a more iterative process of data assimilation, where forecasts are updated with respect to additional observations, rather than projecting 100s of time steps into the future [@dietze_iterative_2018]. 
Updating a forecast after a critical transition has already occurred may be of little use in the context of hysteresis, such as under the saddle node or stochastic transition -- recognizing the alternative stable state only after the system is stuck in that basin will often be considered 'too late'.
Assimilation may be more applicable to the Hopf bifurcation, where additional observations of slowly growing oscillations may lead to more accurate forecasts. 
Such models may even accurately predict the homoclinic bifurcation that occurs when the limit cycle grows too large, eventually hitting a saddle point of zero population size for the host species.
We leave these cases to future exploration rather than attempting to explore all such variations in a single narrative.
